---
title: "Forcats Package"
author: "David Asare Kumi"
date: "4/7/2020"
output: html_document
---


## Defining Factors

- The forcats package is part of the tidyverse and is useful for dealing with factors. Factors are simply categorical variables, useful for controlling the levels and order of a vector.

- Categorical variable or discrete variables, as opposed to continuous variables, are often qualitiative and can take on a finite number of values.

- Examples of categorical variables: types of fruit, locations, party preference, ethnicity,gender etc.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
colors <- c("red","blue","green","green","red","blue","red","green","blue")
class(colors)

#convert colors to a categorical or factor variable
colors <- as.factor(colors)
print(colors)
class(colors)
str(colors)

#Each color is a level, and underlying each level is an integer associated with that level; red = 3, green = 2, blue = 1. The order of the levels is assigned alphabetically to the integer: b = 1, g = 2, r = 3.

#We can reorder the levels if we want to:
colors <- factor(colors, levels = c("red","blue", "green"))
print(colors)
str(colors)

#Let's compare the summary function for colors as a character vector and as a factor.
colors <- c("red","blue","green","green","red","blue","red","green","blue")

summary(colors)

colors <- factor(colors, levels = c("red","blue", "green"))

summary(colors)

#As factors, each color becomes a distinct group and summarized separately.

```

## The Forcats Package

- As useful as factors are, they can be quite a pain to work with at times. Luckily, forcats has functions that allow you to manipulate factors.

- First, let’s install the package using: install.packages("forcats")

- forcats also comes installed as part of the tidyverse suite of packages. It is not part of the core group of tidyverse packages so we must load it explicitly.


### Function: Factor Recode

- One useful function of forcats is fct_recode. This allows you to change the levels (or name/identity) of a factor. Here’s an example:

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(tidyverse)
library(forcats)
library(dplyr)
library(plotly)
data(airquality)
glimpse(airquality)

#The month column is data type integer, let's change it to a factor 
airquality$Month <- factor(airquality$Month)

levels(airquality$Month)

#Let's rename the months using the fct_recode function. 
airquality$Month <- fct_recode(airquality$Month, May = '5', June = '6', July = '7', Aug = '8', Sept = '9')
glimpse(airquality$Month)

g<-ggplot(airquality, aes(Month, Temp)) +
  geom_boxplot(aes(fill = Month)) +
  ggtitle(label = "Daily Temperatures Aggregated by Month")
ggplotly(g)

```

## The Factor Reverse Function

- If you just want to reverse the order, there’s the fct_rev function. You can even use it in line when defining your aesthetics in ggplot like so:

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
g1<-ggplot(airquality, aes(fct_rev(Month), Temp)) +
  geom_boxplot(aes(fill = Month)) +
  labs(x = "Month") +
  ggtitle(label = "Our plot now has the x-axis in reverse order")
ggplotly(g1)

```

## The Factor Relevel Function

- Another useful function is fct_relevel. This function allows us to change any number of levels to any position.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(ggplot2)
library(plotly)

airquality$Month <- fct_relevel(airquality$Month, 'Sept', 'July', 'May', 'Aug', 'June')

levels(airquality$Month)

# This may not seem useful at first, but when you need to visualize your data in a particular way, the fct_relevel function is extremely useful...

g2<-ggplot(airquality, aes(Month, Temp)) +
  geom_boxplot(aes(fill = Month)) +
  ggtitle(label = "Notice how the order of the level 'Month' has changed")
ggplotly(g2)

```

## The Factor Reorder Function

- And finally, it is often useful to reorder the factor in a way that is useful for visualization. For this, we can use the fct_reorder function.

- For this example, let’s use the mtcars data set:

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(ggplot2)
library(plotly)
library(forcats)
data(mtcars)

mtcars$model <- row.names(mtcars)

glimpse(mtcars)

mtcars$model <- factor(mtcars$model)

g3<-ggplot(mtcars, aes(mpg, model)) +
  geom_point() +
  ggtitle(label = "MPG vs. Car Model")
ggplotly(g3)

```

- It’s difficult to make comparisons when the data is scattered. But we’re in luck! We can use the fct_reorder function to clean it up.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(ggplot2)
library(plotly)
library(forcats)
data(mtcars)

mtcars$model <- row.names(mtcars)

#fct_reorder takes three arguments: f = factor you want to reorder, x = the variable in which the order will be based upon, and optionally fun (a function to  be used if there are multiple values of x for each value of f.) Here we focus on only the first two arguments.

ggplot(mtcars, aes(mpg, fct_reorder( model,mpg))) +
 geom_point() + labs(y = "model") +
 ggtitle(label = "We can make better comparison by reordering the levels based on the mpg values!") +
  theme(plot.title = element_text(size = 10, face = 'bold'))

```




