---
title: "Purrr Package"
author: "David Asare Kumi"
date: "4/7/2020"
output: html_document
---


## Example 

### Method 1

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
#load libraries
library(tidyverse)
library(gapminder)
library(purrr)

pop<-rnorm(10000)
n<-1:1000
m1<-rep(NA,length(n))

#using for loops
for(i in n){
  x<-sample(pop,i)
  m1[i]<-mean(x)
}
plot(m1,type="l")

```

- The sample mean slowly converges to zero(0) as n increases.

### Method 2

- Using purrr which is faster than for loops.

- It's all about the map() function.

- ~ means you want to apply a function.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
#load libraries
library(tidyverse)
library(gapminder)
library(purrr)
library(dplyr)

#x is a list or atomic vector
m2<-map(n,~sample(pop,.x))
head(m2) # This is a list

m2<-map(n,~sample(pop,.x))%>%
  map(~mean(.x))
head(m2) # This gives a list

m2<-map(n,~sample(pop,.x))%>%
  map_dbl(~mean(.x))
head(m2)
plot(m2,type="l")

```

- m2 is a list and we have to transform to a vector before we can plot m2 so we employ map_dbl.

- map_dbl(.x,.f,...) is a numeric vector.

- As sample size increases to 1000 the sample mean approaches zero(0).

- map() without anything, the output is a list i.e. base map() function.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
#load libraries
library(tidyverse)
library(gapminder)
library(purrr)
library(dplyr)

m1<-data.frame(n=numeric(1000),mean=numeric(1000),sd=numeric(1000))
for(i in n){
  x<-sample(pop,i)
  m1[i,]<-list(i,mean(x),sd(x))
}
plot(m1$n,m1$mean,type="l",col="blue")
plot(m1$n,m1$sd,type="l",col="red")

```

## Use Purrr

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
#load libraries
library(tidyverse)
library(gapminder)
library(purrr)
library(dplyr)

m2<-map(n,~sample(pop,.x))%>%
  map_df(~data.frame(mean=mean(.x),sd=sd(.x)))
head(m2)

#you can add .id
m2<-map(n,~sample(pop,.x))%>%
  map_df(~data.frame(mean=mean(.x),sd=sd(.x)),.id="n")
head(m2)

plot(m2$n,m2$sd,type="l",col="blue")
plot(m2$n,m2$mean,type="l",col="red")

```

- The extra column n gives the sample sizes.

- As the sample size increases, the sample standard deviation approaches the population standard deviation of 1 and the sample mean approaches the population mean of zero(0).

## Use a data frame

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
#load libraries
library(tidyverse)
library(gapminder)
library(purrr)
library(dplyr)
data(gapminder)
head(gapminder,15)

#model
f1<-lifeExp~pop
f2<-lifeExp~gdpPercap
f3<-lifeExp~pop+gdpPercap
f4<-lifeExp~pop+gdpPercap+year
f5<-lifeExp~pop+gdpPercap+year+continent

m1<-lm(f1,data=gapminder)
m2<-lm(f2,data=gapminder)
m3<-lm(f3,data=gapminder)
m4<-lm(f4,data=gapminder)
m5<-lm(f5,data=gapminder)

summary(m1)
summary(m2)
summary(m3)
summary(m4)
summary(m5)

#if we want the best model
AIC(m1,m2,m3,m4,m5)

```

- The one with the lowest AIC value is the best model.

## Create a formula of 5 variables

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
#load libraries
library(tidyverse)
library(gapminder)
library(purrr)
library(dplyr)
data(gapminder)

formulas<-list(f1,f2,f3,f4,f5)
formulas #The output has [[]] indicating it is a list
mod<-map(formulas,~lm(.x,data=gapminder))%>%
  map(~summary(.x))
mod

```



```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
#load libraries
library(tidyverse)
library(gapminder)
library(purrr)
library(dplyr)
data(gapminder)


formulas<-list(f1,f2,f3,f4,f5)
formulas #The output has [[]] indicating it is a list
mod<-map(formulas,~lm(.x,data=gapminder))%>%
  map_df(~data.frame(formula=format(formula(.x)),df=df.residual(.x),AIC=AIC(.x),stringsAsFactors=F))
mod

```

## Example

- Run model on each country.

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
gap_bycty<-gapminder%>%
  split(.$country)

head(gap_bycty)

gap_bycty[[1]]

gap_bycty$Zimbabwe

mod_bycty<-map(gap_bycty,~lm(lifeExp~year,data=.x))
head(mod_bycty)

mod_bycty$Afghanistan %>%
  summary()

#Alternatively

mod_bycty<-map(gap_bycty,~lm(lifeExp~year,data=.x))%>%
  map(~summary(.x))%>%
  map(~.x$r.squared)
head(mod_bycty)

```



```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
mod_bycty<-map(gap_bycty,~lm(lifeExp~year,data=.x))%>%
  map(~summary(.x))%>%
  map("r.squared")
head(mod_bycty)

```

## Map functions in purrr(R tidyverse)

```{r,echo=TRUE,message=FALSE,warning=FALSE,comment=""}
library(tidyverse)
my_list<-list(
  c(1,2,6),
  c(4,7,1),
  c(9,1,5)
)
#find the mean of each element
my_list[[1]]%>%mean()
my_list[[2]]%>%mean()
my_list[[3]]%>%mean()

#basic idea of a map(general)-simpler and quicker
my_list%>%map(mean) #OR
map(my_list,mean)

#specific map
my_list%>%map_dbl(mean)

#anonymous function
#my_list%>%map(~ * 2)

#Other maps
my_list%>%map(is.numeric)
my_list%>%map_lgl(is.numeric)
my_list%>%map_int(is.numeric)

```





